Bow shocks, bow waves, and dust waves. III. Diagnostics

### 2019-06-22 papers/Papers1234/app-cloudy-dust-heating.tex
Finish fixes to paper 3 for referee

In terms of the characteristic parameters introduced in
\S~\ref{sec:energy-trapp-vers} the dimensionless radiation field
becomes
\begin{equation}
  \label{eq:U-from-L4-and-Rpc}
  U = 14.7\, L_4\, R_{\text{pc}}^{-2} \ ,
\end{equation}
or, alternatively, it can be expressed in terms of the ambient stream as
\begin{equation}
  \label{eq:U-from-ambient}
  U = 3.01 \, n_\infty \, v_{10}^2 / x^2 \ , 
\end{equation}
where \(x = R_0/R_*\) is given by Paper~I's equation~(12).
% XREF: Paper I \eqref{eq:x-cases}
It can also be related to the radiation parameter \(\Xi\), defined in
Paper~II's equation~(23), as
% XREF: Paper II \eqref{eq:Xi-Prad-over-Pgas}
\begin{equation}
  \label{eq:U-vs-Xi}
  U = 3.82 \, n T_4 \, \Xi \ .
\end{equation}
A common alternative approach to scaling the radiation field (see
\citealp{Tielens:1985a} and citations thereof) is to normalize in the
FUV band (\SIrange{0.0912}{0.24}{\um}), where the local interstellar
value is known as the Habing flux \citep{Habing:1968a}:
\begin{equation}
  \label{eq:Habing-flux}
  F\Hab = \SI{0.0016}{erg.s^{-1}.cm^{-2}} \ .
\end{equation}
The resultant dimensionless flux is often denoted by \(G_0\), and the
relationship between \(G_0\) and \(U\) depends on the fraction
\(f_{\text{fuv}}\) of the stellar luminosity that is emitted in the
FUV band:
\begin{equation}
  \label{eq:G-vs-U}
  G_0 = f_{\text{fuv}} \frac{u\mmp\, c}{F\Hab} \,U = (\text{\numrange{6}{10}}) \,U \ ,
\end{equation}
where we give the range corresponding to early O (\(f_{\text{fuv}} \approx 0.4\)) to early B (\(f_{\text{fuv}} \approx 0.7\)) stars.

>>>>

In terms of the characteristic parameters introduced in
\S~\ref{sec:energy-trapp-vers} the dimensionless radiation field
becomes
\begin{equation}
  \label{eq:U-from-L4-and-Rpc}
  U = 14.7\, L_4\, R_{\text{pc}}^{-2} \ ,
\end{equation}
or, alternatively, it can be expressed in terms of the ambient stream as
\begin{equation}
  \label{eq:U-from-ambient}
  U = 3.01 \, n \, v_{10}^2 / x^2 \ , 
\end{equation}
where \(x = R_0/R_*\) is given by Paper~I's equation~(12).
% XREF: Paper I \eqref{eq:x-cases}
It can also be related to the radiation parameter \(\Xi\), defined in
Paper~II's equation~(23), as
% XREF: Paper II \eqref{eq:Xi-Prad-over-Pgas}
\begin{equation}
  \label{eq:U-vs-Xi}
  U = 3.82 \, n T_4 \, \Xi \ .
\end{equation}
A common alternative approach to scaling the radiation field (see
\citealp{Tielens:1985a} and citations thereof) is to normalize in the
FUV band (\SIrange{0.0912}{0.24}{\um}), where the local interstellar
value is known as the Habing flux \citep{Habing:1968a}:
\begin{equation}
  \label{eq:Habing-flux}
  F\Hab = \SI{0.0016}{erg.s^{-1}.cm^{-2}} \ .
\end{equation}
The resultant dimensionless flux is often denoted by \(G_0\), and the
relationship between \(G_0\) and \(U\) depends on the fraction
\(f_{\text{fuv}}\) of the stellar luminosity that is emitted in the
FUV band:
\begin{equation}
  \label{eq:G-vs-U}
  G_0 = f_{\text{fuv}} \frac{u\mmp\, c}{F\Hab} \,U = (\text{\numrange{6}{10}}) \,U \ ,
\end{equation}
where we give the range corresponding to early O
(\(f_{\text{fuv}} \approx 0.4\)) to early B
(\(f_{\text{fuv}} \approx 0.7\)) stars, calculated from TLUSTY stellar
atmosphere models \citep{Lanz:2003a, Lanz:2007a}.

### 2019-06-22 papers/Papers1234/bs-bw-dw-03.tex
Finish fixes to paper 3 for referee

In Paper~I we derived expressions for the bow radius \(R_0\) in each of
the three well-coupled regimes as a function of the parameters of the
star and the ambient medium.  In Paper~II, we found the criteria for
the dust to decouple from the gas to form a separate dust wave outside
of the hydrodynamic bow shock.  The most important of these is that
the ratio of radiation pressure to gas pressure should exceed a
critical value, \(\Xi_\dag \sim 1000\).

@@

In the main body of the paper, we use \(\tau\) for the radiation pressure
optical depth and \(\tau'\) for the absorption optical depth.  These are
related by the ratio of the corresponding frequency-averaged optical
efficiencies:
\begin{equation}
  \label{eq:tau_tauprime}
  \tau = \frac{Q_P}{Q_{\text{abs}}} \tau' = \left(  1 + \varpi \frac{1 - g}{1 - \varpi}\right) \tau'\ .
\end{equation}
Specific results were given in Figure~6a of Paper~II for an example
grain mixture, designed to reproduce average interstellar medium dust
properties \citep{Weingartner:2001b, Abel:2008a, Ferland:2013a,
  Ferland:2017a}.  For such a mixture the albedo is roughly constant
at \(\varpi \approx 0.5\) throughout the ultraviolet and visible bands,
falling to low values in the infrared.  The asymmetry parameter, or
mean scattering cosine, falls across the ultraviolet band from
\(g \approx 1\) at the highest frequencies to \(g\approx 0.5\) in the FUV.\@ This
yields a conversion factor of
\(Q_P / Q_{\text{abs}} = \text{\numrange{1.2}{1.3}}\) at EUV/FUV
wavelengths.

>>>>

In Paper~I we derived expressions for the bow radius \(R_0\)
(star--apex distance) in each of the three well-coupled regimes as a
function of the parameters of the star and the ambient medium.  In
Paper~II, we found the criteria for the dust to decouple from the gas
to form a separate dust wave outside of the hydrodynamic bow shock.
The most important of these is that the ratio of radiation pressure to
gas pressure should exceed a critical value, \(\Xi_\dag \sim 1000\).

@@

In the main body of the paper, we use \(\tau_P\) for the radiation pressure
optical depth and \(\tau_{\text{abs}}\) for the absorption optical depth.  These are
related by the ratio of the corresponding frequency-averaged optical
efficiencies:
\begin{equation}
  \label{eq:tau_tauprime}
  \tau_P = \frac{Q_P}{Q_{\text{abs}}} \tau_{\text{abs}} = \left(  1 + \varpi \frac{1 - g}{1 - \varpi}\right) \tau_{\text{abs}}\ .
\end{equation}
Specific results were given in Figure~6a of Paper~II for an example
grain mixture, designed to reproduce average interstellar medium dust
properties \citep{Weingartner:2001b, Abel:2008a, Ferland:2013a,
  Ferland:2017a}.  For such a mixture the albedo is roughly constant
at \(\varpi \approx 0.5\) throughout the ultraviolet and visible bands,
falling to low values in the infrared.  The asymmetry parameter, or
mean scattering cosine, falls across the ultraviolet band from
\(g \approx 1\) at the highest frequencies to \(g\approx 0.5\) in the FUV.\@ This
yields a conversion factor of
\(Q_P / Q_{\text{abs}} = \text{\numrange{1.2}{1.3}}\) at EUV/FUV
wavelengths.

### 2019-06-22 papers/Papers1234/sec-observational-diagnostics.tex
Finish fixes to paper 3 for referee

A fundamental parameter is the optical depth, \(\tau\), of the bow
shell to UV radiation, which determines what fraction of the stellar
photon momentum is available to support the shell (see
\S~2.1 of Paper~I).  But the same photons also heat the
dust grains in the bow, which re-radiate that energy predominantly at
mid-infrared wavelengths (roughly \SIrange{10}{100}{\um}) with
luminosity \(L\IR\).  Assuming that Ly\(\alpha\) and mechanical
heating of the dust shell is negligible (see
\S~\ref{sec:unimp-other-heat}) and that the emitting shell subtends a
solid angle \(\Omega\), as seen from the star, then the optical depth
can be estimated as
\begin{equation}
  \label{eq:tau-empirical}
  \tau = -\ln \left( 1 - \frac{4\pi}{\Omega} \frac{L\IR}{L_*} \right)
  \approx \frac{2 L\IR}{L_*} \ ,
\end{equation}
where the last approximate equality holds if \(\tau \ll 1\) and the shell
emission covers one hemisphere.\footnote{%
  Note that the \(\tau\) of Paper~I is not exactly the same as the
  \(\tau\) of equation~\eqref{eq:tau-empirical}, but is larger by a
  factor of
  \(Q_P / Q_{\text{abs}} = 1 + \varpi (1 - g)/(1 - \varpi)\), where
  \(\varpi\) is the grain albedo and \(g\) the scattering asymmetry.
  For standard ISM grain mixtures,
  \(Q_P / Q_{\text{abs}} = \text{\numrange{1.2}{1.3}}\) at EUV/FUV
  wavelengths.}
% This is plotted in Figure~6a of Paper~II for Cloudy's default ISM
% grain mixture (see \S~\ref{sec:grain-temp-emiss} below), where it
% can be seen that
% \(Q_P / Q_{\text{abs}} = \text{\numrange{1.2}{1.3}}\) at EUV/FUV
% wavelengths (\(\lambda = \text{\SIrange{0.05}{0.2}{\um}}\)), so the
% correction is small.

A second important parameter is the thermal plus magnetic pressure in
the shocked shell, which is doubly useful since in a steady state it
is equal to \emph{both} the internal supporting pressure (wind ram
pressure plus absorbed stellar radiation) \emph{and} the external
confining pressure (ram pressure of ambient stream).  The shell
pressure is not given directly by the observations, but can be
determined by the following three steps:
\begin{enumerate}[P1.]
\item \label{P1} The shell mass column (\si{g.cm^{-2}}) can be
  estimated from the optical depth by assuming an effective UV
  opacity: \(\Sigma\shell = \tau / \kappa\)
\item \label{P2} The shell density (\si{g.cm^{-3}}) can be found from
  the mass column if the shell thickness is known:
  \(\rho\shell = \Sigma\shell / h\shell\).  In the absence of other information, a
  fixed fraction of the shell radius can be used.  In particular, we
  normalize by a typical value of one~quarter the star--apex distance:
  \(h_{1/4} = h\shell / (0.25 R_0)\).  This corresponds to a Mach
  number \(\M_0 = \surd 3\) if the stream shock is radiative, or
  \(\M_0 \gg 1\) if non-radiative (see \S~3.2 of Paper~I). Further
  discussion is given in \S~\ref{app:bow-shock-data} below. 
\item \label{P3} Finally, the pressure (\si{dyne.cm^{-2}}) follows by
  assuming values for the sound speed and Alfvén speed:
  \(P\shell = \rho\shell (\sound^2 + \frac12 v\alfven^2) \).
\end{enumerate}
It is natural to normalize this pressure to the stellar radiation
pressure at the shell, so we define a shell momentum efficiency
\newcommand\pc{\ensuremath{_{\text{pc}}}}
\begin{equation}
  \label{eq:eta-shell}
  \eta\shell \equiv \frac{P\shell}{P\rad}
  = \frac{4\pi R_0^2\, (\sound^2 + \frac12 v\alfven^2)\, c\, \tau}{L_*\, \kappa\, h\shell}
  \approx 245 \frac{R\pc \, T_4 \, \tau}{L_4 \, \kappa_{600} \, h_{1/4}} \ , 
\end{equation}
where \(c\) is the speed of light. In the last step we have assumed
ionized gas at temperature \(\num{e4}\,T_4\,\si{K}\) with negligible
magnetic support (\(v\alfven \ll \sound\)) and written the stellar
luminosity and shell parameters in terms of typical values, which we
summarize below.  Note that the shell momentum efficiency is simply
the reciprocal of the radiation parameter from Paper~II's
equation~(23): \(\eta\shell = \Xi\shell^{-1}\), which provides yet a third
use for \(\eta\shell\), since \(\Xi\) is paramount in determining whether
the grains and gas remain well-coupled (see \S~4.4 of Paper~II).

In this section and the remainder of the paper, we employ
dimensionless versions of the stellar bolometric luminosity, \(L_*\),
wind mass-loss rate, \(\dot{M}\), and terminal velocity, \(V\wind\),
together with the ambient stream's mass density, \(\rho\), relative
velocity \(v_\infty\), and effective dust opacity, \(\kappa\).  These are
defined as follows:
\begin{align*}
  % \label{eq:stellar-parameters}
  \dot{M}_{-7} &= \dot{M} / \bigl(\SI{e-7}{M_\odot.yr^{-1}}\bigr) \\
  V_3 &= V\wind / \bigl(\SI{1000}{km.s^{-1}}\bigr) \\
  L_4 &= L_* / \bigl(\SI{e4}{L_\odot}\bigr) \\
  v_{10} &= v_\infty / \bigl( \SI{10}{km.s^{-1}} \bigr) \\
  n &= (\rho / \bar{m}) / \bigl( \SI{1}{cm^{-3}} \bigr) \\
  \kappa_{600} &= \kappa / \bigl( \SI{600}{cm^2.g^{-1}} \bigr) \ ,
\end{align*}
where \(\bar{m}\) is the mean mass per hydrogen nucleon
(\(\bar{m} \approx 1.3 m_{\text{p}} \approx \SI{2.17e-24}{g}\) for solar
abundances).

@@

In Figure~\ref{fig:All-sources-eta-tau} we show the resultant
diagnostic diagram: \(\eta\shell\) versus \(\tau\).  The horizontal axis
shows the fraction of the stellar radiative \emph{energy} that is
reprocessed by the bow shell, while the vertical axis shows the
fraction of stellar radiative \emph{momentum} that is imparted to the
shell, either directly by absorption, or indirectly by the stellar
wind (which is itself radiatively driven).  Radiatively supported bows
(DW, RBW, or RBS cases) should lie on the diagonal line
\(\eta\shell = ( Q_P / Q_{\text{abs}}) \tau \approx 1.25 \tau\), where we have used
the ratio of grain radiation pressure efficiency to absorption
efficiency found in the FUV band for the dust mixture shown in
Paper~II's Figure~6.  Wind-supported bows should lie above this line
and no bows should lie below the \(\eta\shell = \tau\) line, since
\(Q_P\) cannot be smaller than \(Q_{\text{abs}}\).

>>>>

A fundamental parameter is the average optical depth, \(\tau\), of the
bow shell to the stellar radiation. In Appendix~\ref{app:dust-opacity}
we discuss in detail the relation between \(\tau\) and the optical
properties of the grains, as well as the slight differences between
the optical depths for absorption, \(\tau_{\text{abs}}\), scattering,
\(\tau_{\text{scat}}\), and radiation pressure, \(\tau_P\).  In the
single-scattering approximation, a fraction \(1 - e^{-\tau_P}\) of the
stellar photon momentum is available to support the shell (see \S~2.1
of Paper~I).  But the same photons also heat the dust grains in the
bow, which re-radiate that energy predominantly at mid-infrared
wavelengths (roughly \SIrange{10}{100}{\um}) with luminosity \(L\IR\).
Assuming that Ly\(\alpha\) and mechanical heating of the dust shell is
negligible (see \S~\ref{sec:unimp-other-heat}) and that the emitting
shell subtends a solid angle \(\Omega\), as seen from the star, then an
absorption optical depth can be estimated as
\begin{equation}
  \label{eq:tau-empirical}
  \tau_{\text{abs}} = -\ln \left( 1 - \frac{4\pi}{\Omega} \frac{L\IR}{L_*} \right)
  \approx \frac{2 L\IR}{L_*} \ ,
\end{equation}
where the last approximate equality holds if
\(\tau_{\text{abs}} \ll 1\) and the shell emission covers one hemisphere.
The two optical depths, \(\tau_P\) and \(\tau_{\text{abs}}\), are not
exactly the same, see equation~\eqref{eq:tau_tauprime}.  The first
represents radiative momentum transfer, which is governed by the grain
radiation pressure efficiency, \(Q_P\), while the second represents
radiative energy transfer, which is governed by the grain absorption
efficiency, \(Q_{\text{abs}}\).  We adopt
\(\tau_P = 1.25 \tau_{\text{abs}}\) based on Cloudy models of a standard ISM
dust mixture, see Figure~6 of Paper~II and
Appendix~\ref{app:dust-opacity}.  In the remainder of the paper, we
drop the subscript and set \(\tau = \tau_{\text{abs}}\).

A second important parameter is the thermal plus magnetic pressure in
the shocked shell, which is doubly useful since in a steady state it
is equal to \emph{both} the internal supporting pressure (wind ram
pressure plus absorbed stellar radiation) \emph{and} the external
confining pressure (ram pressure of ambient stream).  The shell
pressure is not given directly by the observations, but can be
determined by the following three steps:
\begin{enumerate}[P1.]
\item \label{P1} The shell mass column (\si{g.cm^{-2}}) can be
  estimated from the optical depth by assuming an effective UV
  opacity: \(\Sigma\shell = \tau / \kappa\)
\item \label{P2} The shell density (\si{g.cm^{-3}}) can be found from
  the mass column if the shell thickness is known:
  \(\rho\shell = \Sigma\shell / h\shell\).  In the absence of other information, a
  fixed fraction of the shell radius can be used.  In particular, we
  normalize by a typical value of one~quarter the star--apex distance:
  \(h_{1/4} = h\shell / (0.25 R_0)\).  This corresponds to a Mach
  number \(\M_0 = \surd 3\) if the stream shock is radiative, or
  \(\M_0 \gg 1\) if non-radiative (see \S~3.2 of Paper~I). Further
  discussion is given in \S~\ref{app:bow-shock-data} below. 
\item \label{P3} Finally, the pressure (\si{dyne.cm^{-2}}) follows by
  assuming values for the sound speed and Alfvén speed:
  \(P\shell = \rho\shell (\sound^2 + \frac12 v\alfven^2) \).
\end{enumerate}
It is natural to normalize this pressure to the stellar radiation
pressure at the shell, so we define a shell momentum efficiency
\newcommand\pc{\ensuremath{_{\text{pc}}}}
\begin{equation}
  \label{eq:eta-shell}
  \eta\shell \equiv \frac{P\shell}{P\rad}
  = \frac{4\pi R_0^2\, (\sound^2 + \frac12 v\alfven^2)\, c\, \tau}{L_*\, \kappa\, h\shell}
  \approx 245 \frac{R\pc \, T_4 \, \tau}{L_4 \, \kappa_{600} \, h_{1/4}} \ , 
\end{equation}
where \(c\) is the speed of light. In the last step we have assumed
ionized gas at temperature \(\num{e4}\,T_4\,\si{K}\) with negligible
magnetic support (\(v\alfven \ll \sound\)) and written the stellar
luminosity and shell parameters in terms of typical values, which we
summarize below.  In equation~(23) of Paper~II, we defined a radiation
parameter as the ratio of the radiation pressure to gas pressure,
\(\Xi = P\rad/P\gas\). The value of \(\Xi\) in the shell is therefore
\(\Xi\shell = \eta\shell^{-1}\) if magnetic pressure is negligible, which
provides yet a third use for \(\eta\shell\), since the radiation
parameter is paramount in determining whether the grains and gas
remain well-coupled (see \S~4.4 of Paper~II).

In this section and the remainder of the paper, we employ
dimensionless versions of the star--apex distance, \(R_0\), stellar
bolometric luminosity, \(L_*\), wind mass-loss rate, \(\dot{M}\), and
terminal velocity, \(V\wind\), together with the ambient stream's mass
density, \(\rho\), stream--star relative velocity \(v_\infty\), and effective
dust opacity, \(\kappa\).  These are defined as follows:
\begin{align*}
  % \label{eq:stellar-parameters}
  R_{\text{pc}} &= R_0 / \bigl( \SI{1}{pc} \bigr) \\
  \dot{M}_{-7} &= \dot{M} / \bigl(\SI{e-7}{M_\odot.yr^{-1}}\bigr) \\
  V_3 &= V\wind / \bigl(\SI{1000}{km.s^{-1}}\bigr) \\
  L_4 &= L_* / \bigl(\SI{e4}{L_\odot}\bigr) \\
  v_{10} &= v_\infty / \bigl( \SI{10}{km.s^{-1}} \bigr) \\
  n &= (\rho / \bar{m}) / \bigl( \SI{1}{cm^{-3}} \bigr) \\
  \kappa_{600} &= \kappa / \bigl( \SI{600}{cm^2.g^{-1}} \bigr) \ ,
\end{align*}
where \(\bar{m}\) is the mean mass per hydrogen nucleon
(\(\bar{m} \approx 1.3 m_{\text{p}} \approx \SI{2.17e-24}{g}\) for solar
abundances).

@@

In Figure~\ref{fig:All-sources-eta-tau} we show the resultant
diagnostic diagram: \(\eta\shell\) versus \(\tau\).  The horizontal axis
shows the fraction of the stellar radiative \emph{energy} that is
reprocessed by the bow shell, while the vertical axis shows the
fraction of stellar radiative \emph{momentum} that is imparted to the
shell, either directly by absorption, or indirectly by the stellar
wind (which is itself radiatively driven).  Radiatively supported bows
(DW, RBW, or RBS cases) should lie on the diagonal line
\(\eta\shell = ( Q_P / Q_{\text{abs}}) \tau \approx 1.25 \tau\), see discussion
following equation~\eqref{eq:tau-empirical}.  Wind-supported bows
should lie above this line and no bows should lie below the
\(\eta\shell = \tau\) line, since \(Q_P\) cannot be smaller than
\(Q_{\text{abs}}\).
